Lattices of acyclic pipe dreams

### 2023-05-11 acyclicFacets/draft/nuTamari.tex
Needed more details in the last part of the proof.

\begin{theorem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
All pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.
\end{theorem}

\begin{proof} For the forward direction, assume that $\omega$ is not dominant. 
Let $P_0$ be the greedy pipe dream of $0\omega$. It has no crossing in row 0, and cannot be a partition starting at row 1.
There must be a row $r>1$ with more crossing than in row $r-1$. 
Considering the rightmost column $c$ of $P$ with a crossing in row $r$,  the pipes $r<k$ are crossing at $(c,r)$, $\omega^{-1}(k)<\omega^{-1}(r)$ and are in contact in position $(c+1,r-1)$.
Let $Q$ be the pipe dream obtained from $P_0$ by the flip of the crossing at $(c,r)$ with the contact at $(c+1,r-1)$.
The contact of $Q$ at $(c,r)$ would gives $r\contactLess{Q} k$. In row $0$ we only have contacts and $\omega^{-1}(k)<\omega^{-1}(r)$, combining this  with Lemma~\ref{lem:rectangle} would give  that 
$k \contactLess{Q} r$, a contradiction, hence $Q$ is not acyclic.

For the reverse direction, assume we have a pipe dream $P$ that is not acyclic. The contact graph must contain a cycle $i_1 \to i_2\to\cdots\to i_m\to i_1$.
This is only possible if some pipe $i_j$ and $i_\ell$ with some contact on both side of the crossing.
Hence that crossing is not horizontally nor vertically connected, therefore Lemma~\ref{lem:hv_connected} gives that $\omega$ is not dominant.
\end{proof}

>>>>

\nantel{add in acknowledgement that : Lucas Gagnon had a proof of converse in different context}
\begin{theorem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
All pipe dreams with permutation $0 \omega$ are acyclic if and only if $\omega$ is a dominant permutation.
\end{theorem}

\begin{proof} For the forward implication, assume that $\omega$ is not dominant. 
Let $P_0$ be the greedy pipe dream of $0\omega$. It has no crossing in row 0, and cannot be a partition starting at row 1.
There must be a row $r>1$ with more crossing than in row $r-1$. 
Considering the rightmost column $c$ of $P$ with a crossing in row $r$,  the pipes $r<k$ are crossing at $(c,r)$, $\omega^{-1}(k)<\omega^{-1}(r)$ and are in contact in position $(c+1,r-1)$.
Let $Q$ be the pipe dream obtained from $P_0$ by the flip of the crossing at $(c,r)$ with the contact at $(c+1,r-1)$.
The contact of $Q$ at $(c,r)$ would gives $r\contactLess{Q} k$. In row $0$ we only have contacts and $\omega^{-1}(k)<\omega^{-1}(r)$, combining this  with Lemma~\ref{lem:rectangle} would give  that 
$k \contactLess{Q} r$, a contradiction, hence $Q$ is not acyclic.

\nantel{Here, I went a bit fast before. Here is a much more detailed version}
For the converse, assume we have a pipe dream that is not acyclic. The contact graph must contain a cycle $i_1 \to i_2\to\cdots\to i_m\to i_1$.
The contact $i_1 \to i_2$ is southeast of $i_1$ and the contact  $i_m\to i_1$ northwest of $i_1$. 
Let $i_k$ be the first time where $i_{k-1}\to i_k$ is a contact southeast of $i_1$ and $i_k \to i_{k+1}$\footnote{with the convention that$k+1$ is $1$ if $k=m$} is a contact northwest of $i_1$.
This implies that the pipe $i_k$ cross $i_1$. Consider the case where $i_1<i_k$, hence $i_k$ is the vertical pipe of the crossing. 
The contact $i_k\to i_{j+1}$ must occur after the crossing of $i_1$ and $i_k$, since it is northwest of $i_1$.
This implies that the crossing of $i_1$ and $i_k$ cannot be vertically connected. 
If $k=2$, then the contact of $i_1\to i_2$ is before the crossing, implying that the crossing of $i_1$ and $i_2$ cannot be horizontally connected. 
The Lemma~\ref{lem:hv_connected} gives that $\omega$ is not dominant in this case.
Assume $k>2$ and assume that the crossing $i_1$ and $i_k$ is horizontally connected. The contact $i_1\to i_2$ is southeast of $i_k$ and the contact $i_{k-1}\to i_k$ is northwest.
Therefor there must be a pipe $i_j$ for $1<j<k$, such that the contact $i_{j-1}\to i_j$ is southeast of both $i_1$ and $i_k$, and the contact $i_{j}\to i_{j+1}$ is southeast of $i_1$ but northwest of $i_j$.
This implies that the pipe $i_j$ and $i_k$ cross southeast of $i_1$ and we must have $i_j<i_k$. The crossing of $i_j$ and $i_k$ cannot be horizontally connected since $i_j$ has a contact west of it.
It cannot be vertically connected since $i_k$ has a contact north of it. Again, Lemma~\ref{lem:hv_connected} gives that $\omega$ is not dominant in this case.

The case where $i_1>i_k$ is very similar, reflecting all arguments. Hence, in all cases, if a pipe dream is not acyclic, then the permutation $\omega$ is not dominant. 
\end{proof}
